Wall-to-wall mapping of peat depth from Lidar terrain and airborne radiometrics in Norwegian landscapes

### 2024-12-02 ms/ms.Rmd
ms: minor edits Methods/Results

The performance of DSM must be evaluated with reference to a specific purpose [i.e., map vs. model validation, interpolation vs. extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified mean absolute error (accuracy, original scale), R^2^ (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).

@@

None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and a mean absolute error of \unit{60\,cm}.
For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.

>>>>

The performance of DSM must be evaluated with reference to a specific purpose [i.e., map vs. model validation, interpolation vs. extrapolation, @robertsCrossvalidationStrategiesData2017; @milaNearestNeighbourDistance2022], and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance [@wadouxSpatialCrossvalidationNot2021; @meyerMachineLearningbasedGlobal2022].
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal [@linnenbrinkKNNDMCVKfold2024].
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate [@linnenbrinkKNNDMCVKfold2024].
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation [@wadouxSpatialCrossvalidationNot2021].
We implemented the kNNDM with the *CAST* R package [v.1.0.2, @meyerCASTPackageTraining2024], setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified mean absolute error (error magnitude, original scale), R^2^ (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).

@@

None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \@ref(fig:modelMetrics)).
For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an *R^2^* of 0.34, and a mean absolute error of \unit{60\,cm}.
For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.

### 2024-12-02 ms/ms.tex
ms: minor edits Methods/Results

The performance of DSM must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified mean absolute error (accuracy, original scale), R\textsuperscript{2} (correlation, standardized scale), and Lin's concordance correlation coefficient (accuracy and correlation, standardized scale).

@@

None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and a mean absolute error of \unit{60\,cm}.
For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.

>>>>

The performance of DSM must be evaluated with reference to a specific purpose \citep[i.e., map vs.~model validation, interpolation vs.~extrapolation,][]{robertsCrossvalidationStrategiesData2017, milaNearestNeighbourDistance2022}, and here we aimed to evaluate maps of peat depth across the study areas.
In the absence of additional field work to collect a design-based independent validation set, we used a spatial cross-validation scheme to evaluate model performance \citep{wadouxSpatialCrossvalidationNot2021, meyerMachineLearningbasedGlobal2022}.
Specifically, we used k-Means Nearest Neighbor Distance Matching (kNNDM), which creates cross-validation folds that mimic the spatial prediction task that is defined as the goal \citep{linnenbrinkKNNDMCVKfold2024}.
In particular, kNNDM looks for the spatial assignment of training data to folds that minimizes the difference between two distributions: nearest neighbor distances between training and test locations in the cross-validation, and nearest neighbor distances between training and prediction locations for the model.
That way, the spatial separation between folds is similar to the separation between training and prediction locations -- which increases the quality of the map accuracy estimate \citep{linnenbrinkKNNDMCVKfold2024}.
For spatially clustered training data, this approach strikes a balance between the risk of optimistic metrics from random cross-validation and the risk of pessimistic metrics from other forms of spatial cross-validation \citep{wadouxSpatialCrossvalidationNot2021}.
We implemented the kNNDM with the \emph{CAST} R package \citep[v.1.0.2,][]{meyerCASTPackageTraining2024}, setting prediction locations to all AR5 mire cells in the study area, and choosing a number of folds (5--20) that produced the best match between the two NND distributions.
From the cross-validation we quantified mean absolute error (error magnitude, original scale), R\textsuperscript{2} (explained variation, standardized scale), and Lin's concordance correlation coefficient (error magnitude and explained variation, standardized scale).

@@

None of the models were able to predict peat depth across the study areas with high accuracy (Fig. \ref{fig:modelMetrics}).
For Skrimfjella the best model achieved a concordance correlation coefficient of 0.3, an \emph{R\textsuperscript{2}} of 0.34, and a mean absolute error of \unit{60\,cm}.
For Ørskogfjellet the same values were 0.39, 0.33, and \unit{56\,cm}, so the best model at Ørskogfjellet was slightly more accurate than the best model at Skrimfjella.
These values were derived from kNNDM spatial cross-validation with 20 folds at Skrimfjella and 10 folds at Ørskogfjellet.
